---
title: "Causal Inference for ATM decision making"
author: "Akhil Shah"
date: "May 13, 2016"
fontsize: 10pt
 
output:
  beamer_presentation:
    fig_width: 2
    fig_height: 1
    theme: "AnnArbor"
    colortheme: "dolphin"
    fonttheme: "structurebold"
    keep_tex: true
---

#Every decision-makers wish: 'what if...'
* Consider a situation that will likely require GDP
* Decision makers want to know ``*what if* we don't employ GDP?"
    * and quantify potential outcomes (airborne delay?)
* The challenge: Historical ATM data is *observational*! 
    * not a randomized control trial 
    * counterfactuals are never observed
    * systematic differences in weather/traffic between instances of GDP and not-GDP 
* Causal Inference methods can help estimate *unbiased* counterfactuals 
* The modern methods of Causal Inference employ Machine Learning models
* example: What happens to airborne delay if a GDP is not implemented?

# Causal Inference in a nutshell
* Unbiased counterfactual estimation requires eliminating confounders
* Eliminate counfounding through propensity scores - probability to recive treatment (GDP) based on covariates (weather,traffic)
* Estimation of propensity score (or log-odds of): Linear Logistic regression 

$$\frac{P(GDP=1|X)}{1-P(GDP=1|X)}=F(X) = \beta X$$

* GBM (using K trees) is often better (more flexible and robust) than linear logistic regression

$$F(X) = \sum^K\beta_k g_k(X)$$ 

#A first application: Backdated GDP events

* Recall motivation: a situation (time window) where GDP is to be applied (e.g. inclimate weather)
* Decision making question: "What if we don't implement a GDP?" 
* Staticians call this estimation of "average treatment effect on the treated (ATT)": estimate the counterfactual of not applying a GDP
* What outcome do we want to analyze as a result of applying or not-applying GDP?
    - Airborne delay
* GDP events in response to current conditions may be easier to interpret
    - Less forecast uncertainties than planned GDP

#GDP events: planned vs "backdated"

```{r,echo=FALSE,eval=FALSE}
#------FOR REF ONLY: USED TO CREATE DATSET FOR LATER ANALYSIS------
#from airborneDelay.r
setwd("~/NoBackup/code/nasa/src/causalInfFAA/data")
ap<-"JFK"
airport<-paste(ap,"/ZNY",sep="")

advy.path<-"~/NoBackup/code/nasa/data/advy_2010_2014"

filelist<-dir(advy.path,pattern="*.csv")
filepaths<-unlist(lapply(filelist,function(x){file.path(advy.path,x)}))

#below gdp.ewr is badly named!!! can be any ap defined above
#also need to source twangDataCreation.r or copy and paste getGDP func from it
source("~/NoBackup/code/nasa/src/causalInfFAA/getGDP.r")
gdp.ewr.list<-lapply(c(1:20),function(x){data.frame(getGDP(filepaths[x],airport))})
gdp.ewr<-do.call(rbind,gdp.ewr.list)

gdponly<-gdp.ewr[gdp.ewr$status=="GDP",]
save(gdponly,file="~/NoBackup/code/nasa/src/causalInfFAA/data/JFKGDPOnlyEvents")
```


- JFK 2010-2014: 496 GDP root advisories; 71 were 'backdated' 
- Large variance in time horizons for planned GDP 
```{r gdp_event_types,warning=FALSE,echo=FALSE,fig.height=3,fig.width=4,fig.align='center'}
load(file="~/NoBackup/code/nasa/src/causalInfFAA/data/JFKGDPOnlyEvents")
#lets select only root adv*
gdponlyroot<-gdponly[gdponly$rootAdv==1,]
gdponlyroot$backdated<-as.factor(gdponlyroot$startTime-gdponlyroot$sendTime<0)
#there were 71 backdated and 425 planned GDP events between 2010-2014
#some histogram plotting of how timely these events can be
library(ggplot2)
rootTimeliness<-ggplot(gdponlyroot,aes(as.numeric(startTime-sendTime)))
rootTimeliness + xlim(c(-100,500)) + geom_histogram(binwidth=10) + facet_grid(backdated ~ .) +  labs(title="Distribution of all GDP Root Advisories",x="Start-Send time in Minutes")
```


#Backdated GDP start almost immediately
* Backdated GDP events usually start within 10 minutes prior to sending
```{r gdp_bd_event_dist,warning=FALSE,echo=FALSE,fig.height=3,fig.width=4,fig.align='center'}
#zoom into backdated histogram
bd<-ggplot(gdponlyroot,aes(as.numeric(startTime-sendTime),backdated=TRUE))
bd + xlim(c(-40,0)) + geom_histogram(binwidth=1) + labs(title="Distribution of BackDated GDP Root Advisories",x="Start-Send time in Minutes")
```

#A note about data sets and alignment
* Analysis requires hourly TAF, ASPM (various modules), TFMI  
* ASPM has hourly values for average airborne delay
    * Queue length from TAER module of ASPM is 15-min based
    * Arrival Demand and Effective Arrivals (number of serviced aircraft)
* TAF is (nominally) generated for 0,6,12,18h 
* TFMI is event based 
    * GDP initations (root advisories) can be modified
    * Must follow the sub time-series to get actual start/stop times
    * End-result is 'status' of each hour (GDP or No-GDP)

#Airborne delay as an outcome

```{r abdelay_hist,warning=FALSE, echo=FALSE,fig.height=3,fig.width=4,fig.align='center'}
load("~/NoBackup/code/nasa/src/causalInfFAA/data/JFK_backdatedGDPOnly_twang_withASPMweather_noNA_timestamped_data.Rdata")
#show histogram difference between qlength when gdp and not
abdelay_hist<-ggplot(twang.complete.bd,aes(AirDelay_JFK)) + geom_histogram(binwidth=1) +  xlim(c(-20,20)) +facet_grid(status ~ .,scales = "free_y",labeller = label_both)
abdelay_hist+labs(title="Hourly Avg. Airborne Delay and GDP status")
```


#Arrival Queue Length - Predictor or Outcome?
```{r queue_length,warning=FALSE, echo=FALSE,fig.height=3,fig.width=4,fig.align='center'}

#show histogram difference between qlength when gdp and not
qlen_hist<-ggplot(twang.complete.bd,aes(qlength)) + geom_histogram(binwidth=5) +facet_grid(status ~ .,scales = "free_y",labeller = label_both)
qlen_hist+labs(title="Arrival Queue Length at JFK based on GDP status")
```

#A possbily relevant feature: Number of Arrivals 
```{r num_arrivals,warning=FALSE, echo=FALSE,fig.height=3,fig.width=4,fig.align='center'}
#show histogram difference between Num arrivals when gdp and not
ggplot(twang.complete.bd,aes(A_JFK)) + geom_histogram(binwidth=1) +facet_grid(status ~ .,scales = "free_y",labeller = label_both) + labs(title="Number of Arrivals at JFK based on GDP status")
```

#Relevant Features
Strongest predictor of GDP is Queue Length, followed by number of arrivals
```{r feat_rel,warning=FALSE, echo=FALSE,fig.height=3,fig.width=5,fig.align='center'}
#do some feature relevance analysis with xgboost
library(xgboost)
load("~/NoBackup/code/nasa/src/causalInfFAA/data/JFK_twang_withASPMweather_noNA_timestamped_data.Rdata")
#FIRST include qlength as a feature
#get rid of timestamp; hours; AirDelay; status
xgbdat<-twang.complete[,c(-1,-9,-11,-13)] 
dtrain <- xgb.DMatrix(data = as.matrix(xgbdat),label=twang.complete$status)
#create the model
model <- xgboost(data = dtrain, nrounds = 20, objective = "binary:logistic",verbose=0)
imp_matrix <- xgb.importance(feature_names = colnames(xgbdat), model = model)
xgb.plot.importance(importance_matrix = imp_matrix)

```

#Relevant Features 2
```{r feat_rel_noQueue,warning=FALSE, echo=FALSE, fig.height=3,fig.width=5,fig.align='center'}
#NOW get rid of qlength and recalculate feature importance
xgbdat<-twang.complete[,c(-1,-9,-11,-12,-13)] 
dtrain <- xgb.DMatrix(data = as.matrix(xgbdat),label=twang.complete$status)
#create the model
model <- xgboost(data = dtrain, nrounds = 20, objective = "binary:logistic",verbose=0)
#generate importance matrix
imp_matrix <- xgb.importance(feature_names = colnames(xgbdat), model = model)
xgb.plot.importance(importance_matrix = imp_matrix)
```


