---
title: "Causal Inference for ATM decision making"
author: "Akhil Shah"
date: "May 13, 2016"
 
output:
  beamer_presentation:
    fig_width: 4
    fig_height: 3
---
Counterfactual Estimation
========================================================
* Consider a situation that will likely require GDP
* Decision makers would like to quantify *what if* we don't apply a GDP
* Causal Inference methods can help estimate this counterfactual
* Use machine learning methods to generate these counterfactuals
* example: What happens to airborne delay if a GDP is not implemented?

Causal Inference in a nutshell
========================================================
* Unbiased counterfactual estimation requires eliminating confounders
* Eliminate counfounders through propensity scores
* Estimate propensity score through ML: Gradient Boosting Machines (GBM)
* GBM more flexible and robust that logistic regression modeling 

GDP events: planned vs "backdated"
========================================================
```{r,echo=FALSE,eval=FALSE}
#from airborneDelay.r
setwd("~/NoBackup/code/nasa/src/causalInfFAA/data")
ap<-"JFK"
airport<-paste(ap,"/ZNY",sep="")

advy.path<-"~/NoBackup/code/nasa/data/advy_2010_2014"

filelist<-dir(advy.path,pattern="*.csv")
filepaths<-unlist(lapply(filelist,function(x){file.path(advy.path,x)}))

#below gdp.ewr is badly named!!! can be any ap defined above
#also need to source twangDataCreation.r or copy and paste getGDP func from it
source("~/NoBackup/code/nasa/src/causalInfFAA/getGDP.r")
gdp.ewr.list<-lapply(c(1:20),function(x){data.frame(getGDP(filepaths[x],airport))})
gdp.ewr<-do.call(rbind,gdp.ewr.list)

gdponly<-gdp.ewr[gdp.ewr$status=="GDP",]
save(gdponly,file="~/NoBackup/code/nasa/src/causalInfFAA/data/JFKGDPOnlyEvents")
```
JFK 2010-2014: 496 GDP root advisories; 71 were 'backdated' 
* Large variance in time horizons for planned GDP, 
```{r,echo=FALSE,fig.height=3,fig.width=4,fig.align='center'}
load(file="~/NoBackup/code/nasa/src/causalInfFAA/data/JFKGDPOnlyEvents")
#lets select only root adv*
gdponlyroot<-gdponly[gdponly$rootAdv==1,]
gdponlyroot$backdated<-as.factor(gdponlyroot$startTime-gdponlyroot$sendTime<0)
#there were 71 backdated and 425 planned GDP events between 2010-2014
#some histogram plotting of how timely these events can be
library(ggplot2)
rootTimeliness<-ggplot(gdponlyroot,aes(as.numeric(startTime-sendTime)))
rootTimeliness + xlim(c(-100,500)) + geom_histogram(binwidth=10) + facet_grid(backdated ~ .) +  labs(title="Distribution of all GDP Root Advisories",x="Start-Send time in Minutes")
```

* whereas backdated GDP events usually start within 10 minutes prior to sending
```{r,echo=FALSE,fig.height=3,fig.width=4,fig.align='center'}
#zoom into backdated histogram
bd<-ggplot(gdponlyroot,aes(as.numeric(startTime-sendTime),backdated=TRUE))
bd + xlim(c(-40,0)) + geom_histogram(binwidth=1) + labs(title="Distribution of BackDated GDP Root Advisories",x="Start-Send time in Minutes")
```

Airborne Delay and Arrival Queue dynamics
========================================================
* What effect does a GDP have on these?
* Consider +/-5 hours centered on backdated GDP events
```{r,echo=FALSE,eval=FALSE}
gdp.backdated<-gdponlyroot[gdponlyroot$startTime-gdponlyroot$sendTime<0,]
#get rid of that bulky full text advisory
gdp.bd<-gdp.backdated[,-8]
#round to nearest hour then convert to nY time*
gdp.bd$startHour<-round(gdp.bd$startTime,"hour") 
#after formatting, the result is a character*
gdp.bd$NYstartHour<-format(as.POSIXct(gdp.bd$startHour),tz="America/New_York")
save(gdp.bd,file="~/NoBackup/code/nasa/src/causalInfFAA/data/JFKGDPBackDatedOnlyEvents")
```


```{r,echo=FALSE,eval=FALSE}
load("~/NoBackup/code/nasa/src/causalInfFAA/data/JFKGDPBackDatedOnlyEvents")
load("~/NoBackup/code/nasa/src/causalInfFAA/data/JFK_ASPMQueueLength_timestamped_data.Rdata")
#to extract aspm index compare correct tz*
#we're going to cycle thru each gdp.bd start hour in NY time zone
#and find the the aspm index which corresponds to it by matchgin the 
#aspm timestamp
load("~/NoBackup/code/nasa/src/causalInfFAA/data/JFK_ASPMAirborneDelay_timestamped_data.Rdata")
#fix the column name for air delay
colnames(aspm)<-c("timestamp","A","AirDelay")
aspm_time_idx<-unlist(lapply(seq(1, length(gdp.bd$NYstartHour)),function(x){
	which(aspm$timestamp==as.POSIXlt(gdp.bd$NYstartHour[x],tz="America/New_York"))}))
qlength_time_idx<-unlist(lapply(seq(1, length(gdp.bd$NYstartHour)),function(x){
	which(qdat.hourly$timestamp==as.POSIXlt(gdp.bd$NYstartHour[x],tz="America/New_York"))}))
#do pm 5 hours from start of backdated GDP
pm=5
airdelay<-lapply(seq(1:length(aspm_time_idx)), function(x){
	aspm[seq(aspm_time_idx[x]-pm,aspm_time_idx[x]+pm),]$AirDelay})

qlength<-lapply(seq(1:length(qlength_time_idx)), function(x){
	qdat.hourly[seq(qlength_time_idx[x]-pm,qlength_time_idx[x]+pm),]$qlength})
#now melt these for easy plotting: L1 is the event id (recall there are 71 backdated gdp events)
#the time variable ranges from -5 to +5, where time=0 is start of backdated gdp event
library(reshape)
ad<-melt.list(airdelay)
ad$time<-rep(seq(-pm,pm),length(airdelay))
#qlength data from Amy (in qdat.hourly) only goes until 8/31/2014 whereas
#aspm data goes until 12/31/2014.  Thus only 66 backdated events in ql
ql<-melt.list(qlength)
ql$time<-rep(seq(-pm,pm),length(qlength))
save(ad,file="~/NoBackup/code/nasa/src/causalInfFAA/data/AirDelayDynamicsPM5HoursAroundBackDatedGDP")
save(ql,file="~/NoBackup/code/nasa/src/causalInfFAA/data/QueueLengthDynamicsPM5HoursAroundBackDatedGDP")
```

```{r,echo=FALSE,fig.width=6,fig.height=5}
load("~/NoBackup/code/nasa/src/causalInfFAA/data/AirDelayDynamicsPM5HoursAroundBackDatedGDP")
load("~/NoBackup/code/nasa/src/causalInfFAA/data/QueueLengthDynamicsPM5HoursAroundBackDatedGDP")
library(ggplot2)

#plot the first 4 events for queue length and airborne delay dynamics
ggplot(ql[ql$L1<5,],aes(x=time,y=value,group=L1))+geom_smooth(alpha=0.01, size=.5,aes(colour=factor(L1))) + geom_point(aes(colour=factor(L1)))+labs(title="Dynamics of Arrival Queue Around a backdated GDP",x="Relative time [hours]",y="Number in Arrival Queue")
ggplot(ad[ad$L1<5,],aes(x=time,y=value,group=L1))+geom_smooth(alpha=0.01, size=.5,aes(colour=factor(L1))) + geom_point(aes(colour=factor(L1)))+labs(title="Dynamics of Airborne Delay Around a backdated GDP",x="Relative time [hours]",y="Airborne Delay [minutes]")

```

Data manipulation and alignment
=======================================================
* Analysis requires hourly TAF, ASPM (various modules), TFMI  
* ASPM has hourly values for average airborne delay
  * Queue length from TAER module of ASPM is 15-min based
  * 
* TAF is (nominally) generated for 0,6,12,18h
* TFMI is event based 
  * GDP initations (root advisories) can be modified
  * Must follow the sub time-series to get start/stop times



Feature Relevance
=======================================================

```{r, echo=FALSE}
#do some feature relevance analysis with xgboost
library(xgboost)
#get rid of timestamp; hours; AirDelay; status
xgbdat<-twang.complete[,c(-1,-9,-11,-13)] 
dtrain <- xgb.DMatrix(data = as.matrix(xgbdat),label=twang.complete$status)
```

Covariate Balancing with PS
=======================================================

```{r, echo=FALSE}
#load dataset for JFK: covs + outcome (air delay) + treatment status (status)
library(twang)
load("JFK_twang_withASPMweather_noNA_timestamped_data.Rdata")

```
